---
title: "CapstoneNHLProposalGraphs"
author: "Drianna Walker"
date: "2026-02-20"
output: html_document
---

```{r}
# packages
library(readr)
library(glmnet)
library(stats)
library(ggplot2)
library(ggrepel)
library(mgcv)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  autodep = TRUE,
  fig.width = 6.4,
  fig.height = 3.6,
  fig.align = "center",
  out.width = "90%"
)
```

```{r}
# resolve local NHL CSV path (env var first)
csv_candidates <- c(
  Sys.getenv("NHL_CSV_PATH", unset = ""),
  "analysis/nhl/data/shots_2007-2024.csv",
  "../analysis/nhl/data/shots_2007-2024.csv",
  "../../analysis/nhl/data/shots_2007-2024.csv",
  "data/shots_2007-2024.csv"
)
csv_candidates <- csv_candidates[nzchar(csv_candidates)]
nhl_csv_path <- csv_candidates[file.exists(csv_candidates)][1]

if (is.na(nhl_csv_path)) {
  stop(
    "NHL CSV not found. Set NHL_CSV_PATH or place shots_2007-2024.csv at analysis/nhl/data/shots_2007-2024.csv or ../analysis/nhl/data/shots_2007-2024.csv",
    call. = FALSE
  )
}
```

```{r, cache=TRUE, cache.path=file.path("..", "..", ".knitr-cache", "nhl-proposal-models-load/"), cache.extra=file.info(nhl_csv_path)$mtime}
# load data
nhl.data <-read_csv(nhl_csv_path, show_col_types = FALSE)
# rid data of incomplete cases
nhl.data <-nhl.data[complete.cases(nhl.data), ]
```

  The scatterplot for Figure 1 displays the relationship between Shot Difficulty Index (SDI) and actual goal percentage for NHL players with at least 50 shots. Each point represents a player, with point size indicating shot volume color representing the residual, actual minus expected goal percentage. The dashed blue line shows the linear trend, while labeled points highlight extreme performers: players with the highest/lowest residuals, highest SDI, and lowest goal percentages. The negative slope confirms that higher difficulty shots result in lower goal percentages, with overperformers (green) and underperformers (red) identifies relative to expectations.
  In Figure 2, the scatterplot examines the relationship between total shot volume and goal percentage for NHL players with at least 50 shots. Each point represents a player, with the dashed red line showing the linear trend. Labeled points identify the top five players in goal percentage, bottom five players in goal percentage, and top five in shot volume. The trend line leaning relatively horizontal suggests that higher volume shooters do not necessarily have higher efficiency, highlighting the trade-off between quantity and quality of shots. Players such as Sam Colangelo demonstrate elite efficiency despite lower volume, while Sidney Crosby maintains solid percentages on high volume.
  The generalized additive model (GAM) plot in Figure 3 illustrates the marginal effect of shot distance on goal probability, controlling for contributing factors. The y-axis shows the log-odds contribution relative to the median shot distance, with the blue line representing the estimated smooth function and the shaded ribbon indicating 95% confidence intervals. The red dotted line marks the median distance at x=33.5, while the grey dashed line at y=0 represents the baseline. Shots taken from closer than 33.5 feet show positive log-odds contributions, meaning they have higher than average scoring probability, while shots beyond 33.5 feet show increasingly negative contributions. The steep overall negative trend confirms that as shot distance increases, the probability of scoring decreases dramatically, with shots beyond 60 feet showing significantly worse outcomes compared to the average shot. The slight increase in probability from 80-90ft is most likely due to when the goalie is pulled for an extra attacker, leaving shots from across the ice at a higher success rate, or the goalie being caught out of position.


```{r}
# normalize shot distance 
max_dist <-max(nhl.data$shotDistance, na.rm = TRUE)
nhl.data$difficulty_distance <- (nhl.data$shotDistance / max_dist)*100

# Shot angle difficulty
max_angle <-max(abs(nhl.data$shotAngle), na.rm = TRUE)
nhl.data$difficulty_angle <- (abs(nhl.data$shotAngle) / max_angle)*100

# account for diffulty of rebound shots
nhl.data$difficulty_rebound <-ifelse(nhl.data$shotRebound == 1, 30, 0)

# account for difficulty of goalie froze
nhl.data$difficulty_goalie_froze <-ifelse(nhl.data$shotGoalieFroze == 1, 20, 0)

# Combine into SDI (0-100 scale)
nhl.data$SDI <- (nhl.data$difficulty_distance * 0.4+ 
                  nhl.data$difficulty_angle * 0.3+ 
                  nhl.data$difficulty_rebound * 0.2+ 
                  nhl.data$difficulty_goalie_froze *0.1)

# Aggregate by player
player_stats <-aggregate(cbind(goal, SDI, xGoal) ~ shooterName, 
                         data =nhl.data, 
                         FUN = function(x) c(mean=mean(x), n= length(x)))

# clean up the aggregated data
player_stats <-do.call(data.frame, player_stats)
colnames(player_stats) <-c("shooterName", "goal_mean", "shots", 
                           "SDI_mean", "SDI_n", "xGoal_mean", "xGoal_n")

# residual
player_stats$residual <-player_stats$goal_mean - player_stats$xGoal_mean

# Filter for min shots
min_shots <-100
player_stats <-player_stats[player_stats$shots >= min_shots, ]

# Identify extreme players to label 
top_5_residual <-head(player_stats[order(player_stats$residual, decreasing = TRUE), ], 5)
bottom_5_residual <-tail(player_stats[order(player_stats$residual, decreasing = TRUE), ], 5)
top_5_SDI <-head(player_stats[order(player_stats$SDI_mean, decreasing = TRUE), ], 5)
bottom_5_goal <-head(player_stats[order(player_stats$goal_mean, decreasing = FALSE), ], 5)

# Combine unique extreme players
label_players <- unique(rbind(top_5_residual, bottom_5_residual, 
                              top_5_SDI, bottom_5_goal))

# Create base plot
p <-ggplot(player_stats, aes(x = SDI_mean, y = goal_mean)) +
  geom_point(aes(size = shots, color = residual), alpha = 0.6) +
  scale_size_continuous(range = c(1, 4), name = "Shot Volume") +  # Smaller size range
  scale_color_gradient2(low = "darkred", mid = "gray90", high = "darkgreen",
                       midpoint = 0, name = "FG% Residual") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Shot Difficulty vs Actual Goal Percentage",
    subtitle = "Players sized by volume, colored by FG% residual",
    x = "Shot Difficulty Index (SDI)",
    y = "Actual Goal %"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "right"
  )



# Identify most extreme players 
top_3_residual <-head(player_stats[order(player_stats$residual, decreasing= TRUE), ], 3)
bottom_3_residual <-tail(player_stats[order(player_stats$residual, decreasing= TRUE), ], 3)
top_3_SDI <-head(player_stats[order(player_stats$SDI_mean, decreasing= TRUE), ], 3)
bottom_3_goal <-head(player_stats[order(player_stats$goal_mean, decreasing= FALSE), ], 3)

# Combine extreme players
label_players <- unique(rbind(top_3_residual, bottom_3_residual, 
                              top_3_SDI, bottom_3_goal))

# Create base plot
p <- ggplot(player_stats, aes(x=SDI_mean, y=goal_mean))+
  geom_point(aes(size = shots, color = residual), alpha = 0.6)+
  scale_size_continuous(range = c(1, 4), name = "Shot Volume")+
  scale_color_gradient2(low = "darkred", mid = "grey90", high = "darkgreen",
                       midpoint= 0, name = "FG% Residual")+
  geom_smooth(method = "lm", se= FALSE, color = "blue", linetype = "dashed", alpha = 0.5)+
  labs(
    title = "NHL Shot Difficulty vs Actual Goal Percentage",
    subtitle = "Players sized by volume, colored by FG% residual",
    x= "Shot Difficulty Index (SDI)",
    y= "Actual Goal %",
    caption= "Figure 1"
  )+
  theme_minimal()+
  theme(
    plot.title = element_text(hjust= 0.5, face = "bold"),
    plot.subtitle = element_text(hjust= 0.5),
    legend.position = "right"
  )

# labels for extreme players
p + geom_text_repel(data = label_players,
                    aes(label = shooterName), 
                    size = 3,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    force = 3,
                    max.overlaps = 50,
                    min.segment.length = 0.2,
                    segment.color = "grey50",
                    segment.alpha = 0.5)
```

```{r}
# calculate shot and goal percentage by player
player_stats <-aggregate(goal ~ shooterName, 
                         data = nhl.data, 
                         FUN = function(x) c(sum = sum(x), mean = mean(x), n = length(x)))

# Clean up
player_stats <-do.call(data.frame, player_stats)
colnames(player_stats) <-c("shooterName", "goals", "goal_pct", "total_shots")

# Calculate goal percentage
player_stats$goal_pct <-player_stats$goal_pct *100

# Filter for min shots
min_shots <- 100
player_stats <-player_stats[player_stats$total_shots >= min_shots, ]

# highest goal %
top_goal_pct <-head(player_stats[order(player_stats$goal_pct, decreasing = TRUE), ], 5)
# lowest goal %
bottom_goal_pct <-head(player_stats[order(player_stats$goal_pct, decreasing = FALSE), ], 5)
# Highest volume shooters
top_volume <-head(player_stats[order(player_stats$total_shots, decreasing = TRUE), ], 5)

# Combine above
label_players <-unique(rbind(top_goal_pct, bottom_goal_pct, top_volume))


ggplot(player_stats, aes(x= total_shots, y= goal_pct))+
  geom_point(alpha = 0.5, color = "steelblue3", size = 2)+
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed")+
  geom_text_repel(data = label_players,
                  aes(label = shooterName),
                  size = 3,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  force = 2,
                  max.overlaps = 20,
                  segment.color = "grey50",
                  segment.alpha = 0.5) +
  labs(
    title = "Shot Volume vs Goal Percentage",
    subtitle = "Each point represents a player with minimum 50 shots",
    x = "Number of Shots",
    y = "Goal Percentage (%)",
    caption= "Figure 2"
  )+
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
# GAM model- shot success based on distance
gam_model <-gam(goal ~ s(shotDistance, bs = "tp"), 
                 data = nhl.data, 
                 family = binomial(),
                 method = "REML")

# prediction data for distance effect
distance_range <-data.frame(shotDistance = seq(0, 100, length.out = 200))

# Get predictions on link scale 
predictions <-predict(gam_model, newdata = distance_range, type = "link", se.fit= TRUE)

# median distance as baseline
median_distance <-median(nhl.data$shotDistance, na.rm= TRUE)
baseline_pred <-predict(gam_model, newdata = data.frame(shotDistance = median_distance), 
                         type= "link")
baseline_log_odds <-as.numeric(baseline_pred)

# data centered at median distance
plot_data <-data.frame(
  distance = distance_range$shotDistance,
  log_odds = predictions$fit - baseline_log_odds,
  se_lower = predictions$fit - 1.96* predictions$se.fit - baseline_log_odds,
  se_upper = predictions$fit + 1.96* predictions$se.fit - baseline_log_odds
)

ggplot(plot_data, aes(x = distance, y = log_odds))+
  geom_ribbon(aes(ymin = se_lower, ymax = se_upper), alpha = 0.2, fill = "steelblue")+
  geom_line(color = "darkblue", size = 1)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50")+
  #median goal line
  geom_vline(xintercept = median_distance, linetype = "dotted", color = "green4", alpha = 0.5, linewidth= 2)+
  #neutral zone boundaries
  geom_vline(xintercept= 25, linetype="dashed", color="red")+
  geom_vline(xintercept= 75, linetype="dashed", color="red")+
  labs(
    title = "GAM Distance Effect on Shot Success",
    subtitle = paste0("Marginal Log-odds (relative to median distance: ", 
                      round(median_distance, 1), " feet)"),
    x = "Shot Distance (feet)",
    y = "Marginal Log-odds Contribution",
    caption="Figure 3"
  )+
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
  scale_x_continuous(breaks= seq(0, 100, by = 10))
```

