---
title: "Untitled"
author: "Drianna Walker"
date: "2026-02-16"
output: html_document
---

```{r}
# packages
library(readxl)
library(readr)
library(glmnet)
library(stats)
```
```{r}
getwd()
```

```{r}
setwd("C:/Users/drian/OneDrive")
```
```{r}

# dataset
options(timeout = 600)

# Download from MoneyPuck
download.file(
  "https://peter-tanner.com/moneypuck/downloads/shots_2007-2024.zip",
  "moneypuck_data/shots_2007-2024.zip",
  mode = "wb"
)
unzip("moneypuck_data/shots_2007-2024.zip", list = TRUE)

# extract
unzip("moneypuck_data/shots_2007-2024.zip", exdir = "moneypuck_data")

```

```{r}
nhl.data <- read_csv("moneypuck_data/shots_2007-2024.csv", show_col_types = FALSE)
nhl.data <- nhl.data[complete.cases(nhl.data), ]
```

```{r}
# expected goal vs goal scored

# Calculate actual goal rate for each 10-foot distance
distance_seq <- seq(0, max(nhl.data$shotDistance, na.rm = TRUE), by = 10)
nhl.data$distance_group <- floor(nhl.data$shotDistance / 10) * 10

actual_rate <- aggregate(goal ~ distance_group, data = nhl.data, FUN = mean)
expected_rate <- aggregate(xGoal ~ distance_group, data = nhl.data, FUN = mean)

# Combine data
plot_data <- merge(actual_rate, expected_rate, by = "distance_group")
colnames(plot_data) <- c("Distance", "Actual", "Expected")

# Find max probability for scaling
max_prob <- max(plot_data$Actual, plot_data$Expected, na.rm = TRUE)

# bar plot
barplot(height = t(as.matrix(plot_data[, c("Actual", "Expected")])),
        names.arg = plot_data$Distance,
        xlab = "Goal Probability",
        ylab = "Distance from Goal (feet)",
        main = "Scored vs Expected Goals by Distance",
        col = c("steelblue4", "darkred"),
        xlim = c(0,.35),  
        horiz = TRUE,
        las = 1,
        beside = FALSE,
        legend.text = c("Actual Goals", "Expected Goals"),
        args.legend = list(x = "topright")) 
```

```{r}
numeric_cols <- sapply(nhl.data, is.numeric)

predictor_cols <- names(numeric_cols)[numeric_cols]
predictor_cols <- predictor_cols[predictor_cols != "goal"]

X <- as.matrix(nhl.data[, predictor_cols])
y <- nhl.data$goal

# Fit LASSO model
lasso_model <- glmnet(X, y, family = "binomial", alpha = 1)

# Cross-validation to find optimal lambda
cv_lasso <- cv.glmnet(X, y, family = "binomial", alpha = 1)

# Plot CV results
plot(cv_lasso, main = "LASSO Cross-Validation for Goal Prediction")
```
```{r}
nhl.data$residual <- nhl.data$goal - nhl.data$xGoal

# Aggregate by player
player_residuals <- aggregate(residual ~ shooterName, 
                             data = nhl.data, 
                             FUN = function(x) c(mean = mean(x), 
                                                n = length(x)))
player_residuals <- do.call(data.frame, player_residuals)
colnames(player_residuals) <- c("shooterName", "residual_mean", "shots")

# Filter for players with minimum shot attempts
min_shots <- 100
player_residuals <- player_residuals[player_residuals$shots >= min_shots, ]

# Sort and get top/bottom 10
player_residuals <- player_residuals[order(player_residuals$residual_mean, decreasing = TRUE), ]
top_10 <- head(player_residuals, 10)
bottom_10 <- tail(player_residuals, 10)
plot_data <- rbind(top_10, bottom_10)

# Horizontal bar plot
par(mar = c(4, 8, 2, 2))
barplot(plot_data$residual_mean * 100,
        names.arg = plot_data$shooterName,
        horiz = TRUE,
        las = 1,
        xlab = "Residual (Actual - Expected Goal %)",
        ylab = "",
        main = "Top 10 Overperformers & Underperformers vs xG",
        col = ifelse(plot_data$residual_mean > 0, "steelblue", "darkred"),
        xlim = c(min(plot_data$residual_mean * 100) - 2, 
                 max(plot_data$residual_mean * 100) + 2),
        cex.names=0.7)
abline(v = 0, lty = 2, lwd = 1)
legend("topright", 
       legend = c("Overperformer", "Underperformer"),
       fill = c("steelblue", "darkred"),
       cex = 0.8)
```

```{r}
# Get coefficients at optimal lambda
optimal_lambda <- cv_lasso$lambda.min
coefs <- as.matrix(coef(lasso_model, s = optimal_lambda))

# Extract non-zero coefficients
non_zero <- coefs[coefs != 0, ]
non_zero <- non_zero[order(abs(non_zero), decreasing = TRUE)]

# Create data frame of coefficients (excluding intercept)
coef_df <- data.frame(
  variable = names(non_zero[-1]),  # Exclude intercept
  coefficient = non_zero[-1]
)

# Top most important predictors 
top_pred <- coef_df[1:min(20, nrow(coef_df)), ]

all_coefs <- coef_df[order(abs(coef_df$coefficient), decreasing = FALSE), ]
bottom_pred <- all_coefs[1:min(20, nrow(all_coefs)), ]

```


```{r}
#lasso predictors
par(mar = c(5, 12, 4, 2))
barplot(abs(bottom_pred$coefficient),
        names.arg = bottom_pred$variable,
        horiz = TRUE,
        las = 1,
        xlab = "Coefficient Magnitude",
        ylab = "",
        main = "LASSO Predictors for Goal Scoring",
        col = "steelblue4",
        xlim = c(0, 20),
        cex.names = 0.8)
```

```{r}
# Filter to ONLY shots that resulted in goals
goal_shots <- nhl.data[nhl.data$goal == 1, ]

# Analyze event patterns in goal-scoring plays
summary(goal_shots$event)
table(goal_shots$shotType)
table(goal_shots$lastEventCategory)

# Compare to non-goal shots
nogool_shots <- nhl.data[nhl.data$goal == 0, ]

# Compare means of key variables
mean_goal <- colMeans(goal_shots[, c("shotDistance", "shotAngle", "xCord", "yCord")], na.rm = TRUE)
mean_nogoal <- colMeans(nogool_shots[, c("shotDistance", "shotAngle", "xCord", "yCord")], na.rm = TRUE)

comparison <- data.frame(
  variable = names(mean_goal),
  on_goals = mean_goal,
  on_non_goals = mean_nogoal,
  difference = mean_goal - mean_nogoal
)

print(comparison)
```

```{r}
# Calculate Shot Difficulty Index (SDI) 

# normalize shot distance
max_dist <- max(nhl.data$shotDistance, na.rm = TRUE)
nhl.data$difficulty_distance <- (nhl.data$shotDistance / max_dist) * 100

# Shot angle difficulty (wider angle = harder)
max_angle <- max(abs(nhl.data$shotAngle), na.rm = TRUE)
nhl.data$difficulty_angle <-(abs(nhl.data$shotAngle) / max_angle) * 100

# Rebound shots are harder
nhl.data$difficulty_rebound <-ifelse(nhl.data$shotRebound == 1, 30, 0)

# Goalie frozen difficulty
nhl.data$difficulty_goalie_froze <- ifelse(nhl.data$shotGoalieFroze == 1, 20, 0)

# Combine into SDI (0-100 scale)
nhl.data$SDI <- (nhl.data$difficulty_distance * 0.4 + 
                  nhl.data$difficulty_angle * 0.3 + 
                  nhl.data$difficulty_rebound * 0.2 + 
                  nhl.data$difficulty_goalie_froze * 0.1)

# Aggregate by player
player_stats <- aggregate(cbind(goal, SDI, xGoal) ~ shooterName, 
                         data = nhl.data, 
                         FUN = function(x) c(mean = mean(x), n = length(x)))

# Clean up the aggregated data
player_stats <- do.call(data.frame, player_stats)
colnames(player_stats) <- c("shooterName", "goal_mean", "shots", 
                           "SDI_mean", "SDI_n", "xGoal_mean", "xGoal_n")

# Calculate residual
player_stats$residual <- player_stats$goal_mean - player_stats$xGoal_mean

# Filter for minimum shots
min_shots <- 100
player_stats <- player_stats[player_stats$shots >= min_shots, ]

# Create the plot
library(ggplot2)
library(ggrepel)

# Identify ONLY extreme players to label (much smaller set)
top_5_residual <- head(player_stats[order(player_stats$residual, decreasing = TRUE), ], 5)
bottom_5_residual <- tail(player_stats[order(player_stats$residual, decreasing = TRUE), ], 5)
top_5_SDI <- head(player_stats[order(player_stats$SDI_mean, decreasing = TRUE), ], 5)
bottom_5_goal <- head(player_stats[order(player_stats$goal_mean, decreasing = FALSE), ], 5)

# Combine unique extreme players
label_players <- unique(rbind(top_5_residual, bottom_5_residual, 
                              top_5_SDI, bottom_5_goal))

# Create base plot
p <- ggplot(player_stats, aes(x = SDI_mean, y = goal_mean)) +
  geom_point(aes(size = shots, color = residual), alpha = 0.6) +
  scale_size_continuous(range = c(1, 4), name = "Shot Volume") +  # Smaller size range
  scale_color_gradient2(low = "darkred", mid = "gray90", high = "darkgreen",
                       midpoint = 0, name = "FG% Residual") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Shot Difficulty vs Actual Goal Percentage",
    subtitle = "Players sized by volume, colored by FG% residual",
    x = "Shot Difficulty Index (SDI)",
    y = "Actual Goal %"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "right"
  )

# Add labels only for extreme players
p + geom_text_repel(data = label_players,
                    aes(label = shooterName), 
                    size = 3,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    force = 3,
                    max.overlaps = 50,
                    min.segment.length = 0.2,
                    segment.color = "grey50",
                    segment.alpha = 0.5)
```
```{r}
# Identify ONLY the most extreme players (smaller set)
top_3_residual <- head(player_stats[order(player_stats$residual, decreasing = TRUE), ], 3)
bottom_3_residual <- tail(player_stats[order(player_stats$residual, decreasing = TRUE), ], 3)
top_3_SDI <- head(player_stats[order(player_stats$SDI_mean, decreasing = TRUE), ], 3)
bottom_3_goal <- head(player_stats[order(player_stats$goal_mean, decreasing = FALSE), ], 3)

# Combine unique extreme players
label_players <- unique(rbind(top_3_residual, bottom_3_residual, 
                              top_3_SDI, bottom_3_goal))

# Create base plot
p <- ggplot(player_stats, aes(x = SDI_mean, y = goal_mean)) +
  geom_point(aes(size = shots, color = residual), alpha = 0.6) +
  scale_size_continuous(range = c(1, 4), name = "Shot Volume") +
  scale_color_gradient2(low = "darkred", mid = "gray90", high = "darkgreen",
                       midpoint = 0, name = "FG% Residual") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed", alpha = 0.5) +
  labs(
    title = "NHL Shot Difficulty vs Actual Goal Percentage",
    subtitle = "Players sized by volume, colored by FG% residual",
    x = "Shot Difficulty Index (SDI)",
    y = "Actual Goal %"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "right"
  )

# Add labels only for extreme players
p + geom_text_repel(data = label_players,
                    aes(label = shooterName), 
                    size = 3,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    force = 3,
                    max.overlaps = 20,
                    min.segment.length = 0.2,
                    segment.color = "grey50",
                    segment.alpha = 0.5)
```
The equation to determine SDI is: SDI = 0.4DC + 0.3AC + 0.2RC + 0.1GFC, with DC being Distance Component, AC being Angle Component, RC being Rebound Component, and GFC being Goalie Froze Component.
  This scatterplot displays the relationship between Shot Difficulty Index (SDI) and actual goal percentage for NHL players with at least 100 shots. Each point represents a player, with point size indicating shot volume color representing the residual, actual minus expected goal percentage. The dashed blue line shows the linear trend, while labeled points highlight extreme performers: players with the highest/lowest residuals, highest SDI, and lowest goal percentages. The negative slope confirms that higher difficulty shots result in lower goal percentages, with overperformers (green) and underperformers (red) identifies relative to expectations.

```{r}
# Calculate shots on goal and goal percentage by player
player_shot_stats <- aggregate(cbind(shotWasOnGoal, goal) ~ shooterName, 
                              data = nhl.data, 
                              FUN = function(x) c(sum = sum(x), mean = mean(x), n = length(x)))

# Clean up the aggregated data
player_shot_stats <- do.call(data.frame, player_shot_stats)
colnames(player_shot_stats) <- c("shooterName", 
                                 "shots_on_goal_sum", "shots_on_goal_pct", "shots_on_goal_total_shots",
                                 "goals_sum", "goal_pct", "total_shots")

# Calculate goal percentage 
player_shot_stats$goal_percentage <- player_shot_stats$goals_sum / player_shot_stats$total_shots * 100

# Filter for minimum total shots
min_shots <- 100
player_shot_stats <- player_shot_stats[player_shot_stats$total_shots >= min_shots, ]

# Sort by total shots
player_shot_stats <- player_shot_stats[order(player_shot_stats$total_shots, decreasing = TRUE), ]

# Get top 10 most shots and bottom 10 least shots
most_shots <- head(player_shot_stats, 10)
least_shots <- tail(player_shot_stats, 10)

# Add category
most_shots$category <- "Most Shots"
least_shots$category <- "Least Shots"

# Combine for plotting
plot_data <- rbind(most_shots, least_shots)

# Create side-by-side bar plot
library(ggplot2)

ggplot(plot_data, aes(x = reorder(shooterName, total_shots), y = goal_percentage, fill = category)) +
  geom_col() ]+
  geom_text(aes(label = paste0(total_shots, " shots")), 
            hjust = -0.1, size = 3)+
  coord_flip()+
  scale_fill_manual(values = c("Most Shots" = "darkblue", "Least Shots" = "darkred")) +
  labs(
    title = "Goal Percentage: Players with Most vs Least Shot Volume",
    subtitle = "Minimum 100 total shots | Numbers show total shot count",
    x = "Player",
    y = "Goal Percentage (%)",
    fill = "Category"
  )+
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )+
  scale_y_continuous(limits = c(0, max(plot_data$goal_percentage) * 1.2))
```

```{r}
# Calculate shots and goal percentage by player
player_stats <- aggregate(goal ~ shooterName, 
                         data = nhl.data, 
                         FUN = function(x) c(sum = sum(x), mean = mean(x), n = length(x)))

# Clean up
player_stats <- do.call(data.frame, player_stats)
colnames(player_stats) <- c("shooterName", "goals", "goal_pct", "total_shots")

# Calculate goal percentage
player_stats$goal_pct <- player_stats$goal_pct * 100

# Filter for minimum shots
min_shots <- 50
player_stats <- player_stats[player_stats$total_shots >= min_shots, ]

# Identify top and bottom performers
# Top goal scorers (highest goal %)
top_goal_pct <- head(player_stats[order(player_stats$goal_pct, decreasing = TRUE), ], 5)
# Bottom goal scorers (lowest goal %)
bottom_goal_pct <- head(player_stats[order(player_stats$goal_pct, decreasing = FALSE), ], 5)
# Highest volume shooters
top_volume <- head(player_stats[order(player_stats$total_shots, decreasing = TRUE), ], 5)

# Combine for labeling
label_players <- unique(rbind(top_goal_pct, bottom_goal_pct, top_volume))

# Create scatter plot
library(ggplot2)
library(ggrepel)

ggplot(player_stats, aes(x = total_shots, y = goal_pct)) +
  geom_point(alpha = 0.5, color = "steelblue3", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed") +
  geom_text_repel(data = label_players,
                  aes(label = shooterName),
                  size = 3,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  force = 2,
                  max.overlaps = 20,
                  segment.color = "grey50",
                  segment.alpha = 0.5) +
  labs(
    title = "Shot Volume vs Goal Percentage",
    subtitle = "Each point represents a player with minimum 50 shots",
    x = "Number of Shots",
    y = "Goal Percentage (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```
The scatterplot examines the relationship between total shot volume and goal percentage for NHL players with at least 100 shots. Each point represents a player, with the dashed red line showing the linear trend. Labeled points identify the top five players in goal percentage, bottom five players in goal percentage, and top five in shot volume. The trend line leaning relatively horizontal suggests that higher volume shooters do not necessarily have higher efficiency, highlighting the trade-off between quantity and quality of shots. Players such as Sam Colangelo demonstrate elite efficiency despite lower volume, while Sidney Crosby maintains solid percentages on high volume.

```{r}
# Calculate shots and goal percentage by player
player_stats <- aggregate(cbind(goal, shotWasOnGoal) ~ shooterName, 
                         data = nhl.data, 
                         FUN = function(x) c(sum = sum(x), mean = mean(x), n = length(x)))

# Clean up
player_stats <- do.call(data.frame, player_stats)
colnames(player_stats) <- c("shooterName", 
                           "goals", "goal_pct", "total_shots",
                           "sog", "sog_pct", "sog_total_shots")

# Calculate goal percentage and SOG percentage
player_stats$goal_pct <- player_stats$goal_pct * 100
player_stats$sog_pct <- player_stats$sog_pct * 100

# Filter for minimum total shots
min_goals <- 1
player_stats <- player_stats[player_stats$goals >= min_goals, ]

# Sort by goals scored
player_stats <- player_stats[order(player_stats$goals, decreasing = TRUE), ]

# Get top 10 most goals and bottom 10 least goals
most_goals <- head(player_stats, 10)
least_goals <- tail(player_stats, 10)

# Add category
most_goals$category <- "Most Goals"
least_goals$category <- "Least Goals"

# Combine for plotting
plot_data <- rbind(most_goals, least_goals)

# Create side-by-side bar plot
library(ggplot2)

ggplot(plot_data, aes(x = reorder(shooterName, goals), y = goal_pct, fill = category)) +
  geom_col() +
  geom_text(aes(label = paste0(goals, " goals (", round(sog_pct, 1), "% SOG)")), 
            hjust = -0.05, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("Most Goals" = "darkblue", "Least Goals" = "darkred")) +
  labs(
    title = "Goal Percentage: Players with Most vs Least Goals Scored",
    subtitle = "Minimum 100 total shots | Labels show goals scored and SOG%",
    x = "Player",
    y = "Goal Percentage (%)",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_y_continuous(limits = c(0, max(plot_data$goal_pct) * 1.3))
```

```{r}
# GAM model- shot success based on distance
gam_model <-gam(goal ~ s(shotDistance, bs= "tp"), 
                 data = nhl.data, 
                 family = binomial(),
                 method = "REML")

# prediction data for distance effect
distance_range <-data.frame(shotDistance = seq(0, 100, length.out = 200))

# Get predictions on link scale 
predictions <-predict(gam_model, newdata = distance_range, type= "link", se.fit= TRUE)

# median distance as baseline
median_distance <-median(nhl.data$shotDistance, na.rm= TRUE)
baseline_pred <-predict(gam_model, newdata = data.frame(shotDistance = median_distance), 
                         type= "link")
baseline_log_odds <-as.numeric(baseline_pred)

# data centered at median distance
plot_data <-data.frame(
  distance = distance_range$shotDistance,
  log_odds = predictions$fit - baseline_log_odds,
  se_lower = predictions$fit - 1.96* predictions$se.fit - baseline_log_odds,
  se_upper = predictions$fit + 1.96* predictions$se.fit - baseline_log_odds
)

ggplot(plot_data, aes(x = distance, y = log_odds))+
  geom_ribbon(aes(ymin = se_lower, ymax = se_upper), alpha = 0.2, fill = "steelblue")+
  geom_line(color = "darkblue", size = 1)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50")+
  #median goal line
  geom_vline(xintercept = median_distance, linetype = "dotted", color = "green4", alpha = 0.5, linewidth= 2)+
  #neutral zone boundaries
  geom_vline(xintercept= 25, linetype="dashed", color="red")+
  geom_vline(xintercept= 75, linetype="dashed", color="red")+
  labs(
    title = "GAM Distance Effect on Shot Success",
    subtitle = paste0("Marginal Log-odds (relative to median distance: ", 
                      round(median_distance, 1), " feet)"),
    x = "Shot Distance (feet)",
    y = "Marginal Log-odds Contribution"
  )+
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
  scale_x_continuous(breaks= seq(0, 100, by = 10))
```
  The generalized additive model (GAM) plot illustrates the marginal effect of shot distance on goal probability, controlling for contributing factors. The y-axis shows the log-odds contribution relative to the median shot distance, with the blue line representing the estimated smooth function and the shaded ribbon indicating 95% confidence intervals. The red dotted line marks the median distance at x=33.5, while the grey dashed line at y=0 represents the baseline. Shots taken from closer than 33.5 feet show positive log-odds contributions, meaning they have higher than average scoring probability, while shots beyond 33.5 feet show increasingly negative contributions. The steep overall negative trend confirms that as shot distance increases, the probability of scoring decreases dramatically, with shots beyond 60 feet showing significantly worse outcomes compared to the average shot. The slight increase in probability from 80-90ft is most likely due to when the goalie is pulled for an extra attacker, leaving shots from across the ice at a higher success rate, or the goalie being caught out of position.
